High Resolution Identification of Protein-DNA Binding Events and Quality Control for ChIP-exo data\vspace*{\fill}

### 2015-11-23 rmarkdown/prelims_paper.Rnw
Added section about software

ChIP-exo (Chromatin Immunopecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
(\cite{exo1}) is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation sequencing
) assay.

@@

@@

@@

Figure \ref{fig:enrich}B shows box plots of the $\mbox{local-NSC}$ on
regions of the three replicates, stratified by the number of reads
mapped exactly to one position for all $\sigma^{70}$ ChIP-exo
experiments; The high stratum is defined by regions consisting of more
than \Sexpr{high} unique positions, the medium stratum for regions
where the number of unique position is on the
(\Sexpr{med},\Sexpr{high}) range and the low stratum in the
(\Sexpr{low},\Sexpr{med}) range. For each stratum and biological
replicate, we sampled \Sexpr{strata} regions whenever possible (in the
opposite case, all the regions in the category were considered) and
calculated the $\mbox{local-NSC}$ for those regions. For all three
replicates, it is shown an increasing trend as the number of unique
position increases, which means that the $\mbox{local-NSC}$ is an
effective indicator of library complexity. Figure \ref{fig:enrich}C
shows the $\mbox{local-NSC}$ coefficient calculated for the same set
of regions, stratified by the number of unique positions in the
regions of ERR336942 using the same criteria as before. In the high
panel it is shown that the ERR336942 sample performs slightly better
than ERR336956 and both outperform ERR336935, this agrees with
\ref{fig:enrich}A, where only in the left-most panel doesn't seem to
be a separation between the low $\mbox{ARC}$ arm and the other.

\begin{figure}[H]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
  \includegraphics[width = .8\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate and C)
    Boxplots of the a set of ERR336942 regions, stratified as in (B)
    with the $\mbox{local-NSC}$ calculated with the different
    biological replicates.}
  \label{fig:enrich}
\end{figure}

@@

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least \Sexpr{de}
fragments into into the ones that overlap with any of the sample's
peaks. The p.value for ERR36935 is \Sexpr{pval[[1]]}, for ERR336942 is
\Sexpr{pval[[2]]} and for ERR336956 is \Sexpr{pval[[3]]}. This shows
that only the middle sample from figure \ref{fig:strand} is the least
imbalance sample from the the three, while the other two still show a
higher imbalance at the \Sexpr{de} level.

@@

\subsubsection{Comparison with ChIP-Seq data using dPeak}
\label{sec:dpeak_analysis}

@@

@@

@@

>>>>

ChIP-exo (Chromatin Immunopecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
(\cite{exo1}) is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation sequencing
 assay.

@@

\subsubsection{dPeak's algorithm initialization strategy}
\label{sec:algo}

dPeak's model likelihood is optimized by the stochastic EM
algorithm. Hence, it is of utmost importance how the parameters are
initialized in order to obtain a global maximum.

Chung et al., 2013 \cite{dpeak} derived an initialization strategy for
the stochastic EM algorithm where each region is initialized by
separate independently and the binding sites locations $\mu_g,
g=1,\cdots,g^*$ are initialized by a uniform partition.

For this work, we considered a new strategy where the all the regions
are ranked by a statistical measure (as the $-\log(p)$ where $p$ is
the empirical p.value of the test $H_0: \mathcal{R} \text{ is
  enriched}$, where $\mathcal{R}$ is a genomic region) or a empirical
quantity (as the number of ChIP tag counts in the region) and use as
initial estimates of the $\delta$ and $\sigma$ parameters the average
of the estimates from the highest ranked $N_{\text{top}}$ peaks.

@@

\subsubsection{Local Normalized Strand cross-correlation}
\label{sec:locNSC}

We considered the creation of a novel QC measure for ChIP-exo called
local normalized SCC or $\mbox{local-NSC}$.

@@

Figure \ref{fig:enrich}B shows box plots of the $\mbox{local-NSC}$ on
regions of the three replicates, stratified by the number of reads
mapped exactly to one position for all $\sigma^{70}$ ChIP-exo
experiments; The high stratum is defined by regions consisting of more
than \Sexpr{high} unique positions, the medium stratum for regions
where the number of unique position is on the
(\Sexpr{med},\Sexpr{high}) range and the low stratum in the
(\Sexpr{low},\Sexpr{med}) range. For each stratum and biological
replicate, we sampled \Sexpr{strata} regions whenever possible (in the
opposite case, all the regions in the category were considered) and
calculated the $\mbox{local-NSC}$ for those regions. For all three
replicates, it is shown an increasing trend as the number of unique
position increases, which means that the $\mbox{local-NSC}$ is an
effective indicator of library complexity. Figure \ref{fig:enrich}C
shows the $\mbox{local-NSC}$ coefficient calculated for the same set
of regions, stratified by the number of unique positions in the
regions of the first replicate using the same criteria as before. In
the high panel it is shown that the first replicate sample performs
slightly better than the second and both outperform the third one,
this agrees with \ref{fig:enrich}A, where only in the left-most panel
doesn't seem to be a separation between the low $\mbox{ARC}$ arm and
the other.

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .7\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
  \includegraphics[width = .7\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate and C)
    Boxplots of the a set of the first replicate regions, stratified
    as in (B) with the $\mbox{local-NSC}$ calculated with the
    different biological replicates.}
  \label{fig:enrich}
\end{figure}

@@

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least \Sexpr{de}
fragments into into the ones that overlap with any of the sample's
peaks. The p.value for the third replicate is \Sexpr{pval[[1]]}, for
first replicate is \Sexpr{pval[[2]]} and for the second is
\Sexpr{pval[[3]]}. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the \Sexpr{de}
level.

@@

\subsection{Comparison with ChIP-Seq data using dPeak}
\label{sec:dpeak_analysis}

@@

\section{Software packages}
\label{sec:software}

For this project several R packages were created or updated:

\begin{itemize}
\item \textbf{dPeak}: We updated the initialization strategy. The
  latest version is currently available from
  \url{http://dongjunchung.github.io/dpeak/}.
\item \textbf{ChIPexoQual}: This package contains the QC pipeline for
  ChIP-exo. The last version is available in
  \url{https://github.com/welch16/ChIPexoQual}.
\item \textbf{Segvis}: The goal of this package is to visualize
  genomic regions by using aligned reads such as in figures
  \ref{fig:exo} or \ref{fig:qcdiagram}.  The latest version is
  available in \url{https://github.com/keleslab/Segvis}.
\item \textbf{ChIPUtils}: This package attempts to gather the most
  commonly used ChIP-Seq QC (as in table \ref{tab:qcbase} or figure
  \ref{fig:scc}) indicators and additional utilities (as figure
  \ref{fig:comp}A) for ChIP-Seq. The latest available version is in
  \url{https://github.com/welch16/ChIPUtils}.
\end{itemize}

@@

\subsection{Analysis of E. Coli's Transcription Initiation Complexes}
\label{sec:ecoli}

@@

\section*{Supplement}
\label{sec:supp}

\subsection*{Additional Enrichment plots for $\sigma^{70}$}
\label{sec:enrichsup}

In figure \ref{fig:enrich2} it is shown the same relationship as in
figure \ref{fig:enrich}, but considering only regions formed where the
reads are being allocated in more than 10 (A) and 30 (B) unique
positions respectively. Both plots show that the vertical arms formed
by regions with low $\mbox{ARC}$ is formed by low complexity regions,
that way suggesting that this segment correspond to the background of
a ChIP-exo experiment.

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page = 4]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \caption{Hexbin plots of $\mbox{ARC}$ vs $\mbox{URCR}$ for each
    region after partitioning the genome. In A and B the regions with
    reads mapped to at most 10 and 30 positions respectively where not
    considered. $\mbox{ARC}$ is defined as the ratio of the nr. of
    reads and the width of a region and $\mbox{URCR}$ is the ratio of
    the number of unique position where the reads are being allocated
    and the number of reads in a region.}
  \label{fig:enrich2}
\end{figure}

### 2015-11-23 rmarkdown/prelims_paper.tex
Added section about software

Figure \ref{fig:enrich}B shows box plots of the $\mbox{local-NSC}$ on
regions of the three replicates, stratified by the number of reads
mapped exactly to one position for all $\sigma^{70}$ ChIP-exo
experiments; The high stratum is defined by regions consisting of more
than 100 unique positions, the medium stratum for regions
where the number of unique position is on the
(50,100) range and the low stratum in the
(20,50) range. For each stratum and biological
replicate, we sampled 400 regions whenever possible (in the
opposite case, all the regions in the category were considered) and
calculated the $\mbox{local-NSC}$ for those regions. For all three
replicates, it is shown an increasing trend as the number of unique
position increases, which means that the $\mbox{local-NSC}$ is an
effective indicator of library complexity. Figure \ref{fig:enrich}C
shows the $\mbox{local-NSC}$ coefficient calculated for the same set
of regions, stratified by the number of unique positions in the
regions of ERR336942 using the same criteria as before. In the high
panel it is shown that the ERR336942 sample performs slightly better
than ERR336956 and both outperform ERR336935, this agrees with
\ref{fig:enrich}A, where only in the left-most panel doesn't seem to
be a separation between the low $\mbox{ARC}$ arm and the other.

\begin{figure}[H]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
  \includegraphics[width = .8\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate and C)
    Boxplots of the a set of ERR336942 regions, stratified as in (B)
    with the $\mbox{local-NSC}$ calculated with the different
    biological replicates.}
  \label{fig:enrich}
\end{figure}

@@

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least 100
fragments into into the ones that overlap with any of the sample's
peaks. The p.value for ERR36935 is 0.1039, for ERR336942 is
0.0064 and for ERR336956 is 0.0871. This shows
that only the middle sample from figure \ref{fig:strand} is the least
imbalance sample from the the three, while the other two still show a
higher imbalance at the 100 level.

>>>>

Figure \ref{fig:enrich}B shows box plots of the $\mbox{local-NSC}$ on
regions of the three replicates, stratified by the number of reads
mapped exactly to one position for all $\sigma^{70}$ ChIP-exo
experiments; The high stratum is defined by regions consisting of more
than 100 unique positions, the medium stratum for regions
where the number of unique position is on the
(50,100) range and the low stratum in the
(20,50) range. For each stratum and biological
replicate, we sampled 400 regions whenever possible (in the
opposite case, all the regions in the category were considered) and
calculated the $\mbox{local-NSC}$ for those regions. For all three
replicates, it is shown an increasing trend as the number of unique
position increases, which means that the $\mbox{local-NSC}$ is an
effective indicator of library complexity. Figure \ref{fig:enrich}C
shows the $\mbox{local-NSC}$ coefficient calculated for the same set
of regions, stratified by the number of unique positions in the
regions of the first replicate using the same criteria as before. In
the high panel it is shown that the first replicate sample performs
slightly better than the second and both outperform the third one,
this agrees with \ref{fig:enrich}A, where only in the left-most panel
doesn't seem to be a separation between the low $\mbox{ARC}$ arm and
the other.

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .7\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
  \includegraphics[width = .7\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate and C)
    Boxplots of the a set of the first replicate regions, stratified
    as in (B) with the $\mbox{local-NSC}$ calculated with the
    different biological replicates.}
  \label{fig:enrich}
\end{figure}

@@

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least 100
fragments into into the ones that overlap with any of the sample's
peaks. The p.value for the third replicate is 0.1039, for
first replicate is 0.0064 and for the second is
0.0871. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the 100
level.
